### 2023-12-31 chapters/background.tex
Revisions

\subsection{Maximum likelihood estimation}
%
VAEs are usually mixed-effect models. Although VAEs use variational Bayes to infer a posterior distribution over latent variables, their parameters are estimated using maximum likelihood. 
That is, we want to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$ to find point estimates. 
Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
Hence, we must approximate the expectation in the ELBO \cref{eq:variational-autoencoder-elbo} and compute its gradients w.r.t. $\phib$ and $\thetab$.

@@

Whereas we arrive at $\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)$ by first deriving the ELBO as a lower bound on the marginal likelihood and then estimating that bound by Monte Carlo sampling, \textcite{burda_importance_2016} instead propose to directly Monte Carlo sample the expectation in the logarithm of the intractable marginal likelihood \cref{eq:marginal-likelihood-with-variational-approximation},
%
\begin{equation}
    \widehat{\mathcal{L}}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \equiv \log \left( \frac{1}{S}\sum_{s=1}^{S} \frac{p_\thetab(\xb|\zb_{s})p(\zb_{s})}{q_\phib(\zb_{s}|\xb)} \right) = \frac{1}{S} \sum_{s=1}^S \log w_\thetab(\xb,\zb_{s}) \enspace ,
\end{equation}
%
where $w_\thetab(\xb,\zb_{s})$ are the importance weights. Note the difference with the estimator in \cref{eq:variational-autoencoder-elbo-sampling-expanded}. 
By taking the expectation of this estimator, we can see that it too is a lower bound on the marginal likelihood,
%
\begin{align}
    \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib)
    &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[ \widehat{\mathcal{L}}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \right] \nonumber \\
    &= \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \frac{1}{S}\sum_{s=1}^{S} \log \left( w_\thetab(\xb,\zb_{s}) \right) \right] \nonumber \\
    &\leq \log \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \frac{1}{S}\sum_{s=1}^{S} w_\thetab(\xb,\zb_{s}) \right] \nonumber \\
    &= \log p(\xb) \enspace .
\end{align}
%
From this and the law of large numbers, it can be shown that for all $S$,
%
\begin{equation}
    \log p(\xb) \geq \mathcal{L}^{\text{IWAE}}_{S+1}(\xb;\thetab,\phib) \geq \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \enspace ,
\end{equation}
%
and $\lim_{S\rightarrow\infty} \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \rightarrow \log p(\xb)$, if $w_\thetab(\xb,\zb_{s})$ are bounded. 
Hence, $\widehat{\mathcal{L}}^{\text{IWAE}}_{S}$ is a biased estimator of $\log p(\xb)$ but with a bias that goes to zero as the number of samples goes to infinity. 
With $S=1$, the estimator is equivalent to the regular ELBO \cref{eq:variational-autoencoder-elbo} and so, it is always as tight, or tighter, than the ELBO \parencite{burda_importance_2016}.

>>>>

\subsection{Maximum likelihood estimation}
%
VAEs are almost always defined as mixed-effect models: VAEs use variational Bayes to infer a posterior distribution over latent variables, but their parameters are estimated using maximum likelihood. 
Hence, parameter estimation entails jointly optimizing the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$ to find point estimates. 
Gradient-based optimization works well for neural networks, but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
Furthermore, we must approximate the expectation over the approximate posterior in the ELBO \cref{eq:variational-autoencoder-elbo} and compute its gradients w.r.t. $\phib$ and $\thetab$. 
We will look into how batching affects the form of the ELBO in \cref{subsec: mutual-information-interpretation-of-elbo} and move on with posterior sampling and gradient estimation here.

@@

In the above we arrived at $\widehat{\mathcal{L}}_{S}(\xb;\thetab,\phib)$ by first deriving the ELBO of \cref{eq:latent-variable-model-elbo} as a lower bound on the marginal likelihood and then approximating that bound by Monte Carlo sampling. Instead, \textcite{burda_importance_2016} propose to Monte Carlo sample the expectation inside the logarithm of \cref{eq:marginal-likelihood-with-variational-approximation} directly, without first deriving the ELBO. They obtain
%
\begin{equation}
    \widehat{\mathcal{L}}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \equiv \log \left( \frac{1}{S}\sum_{s=1}^{S} \frac{p_\thetab(\xb|\zb_{s})p(\zb_{s})}{q_\phib(\zb_{s}|\xb)} \right) = \frac{1}{S} \sum_{s=1}^S \log w_\thetab(\xb,\zb_{s}) \enspace ,
\end{equation}
%
where $w_\thetab(\xb,\zb_{s})$ can be seen as importance weights. Note that the location of the logarithm is the only difference with the estimator in \cref{eq:variational-autoencoder-elbo-sampling-expanded}. 
By taking the expectation of this estimator, we can see that it too is a lower bound on the marginal likelihood,
%
\begin{align}
    \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib)
    &= \mathbb{E}_{q_\phib(\zb|\xb)}\left[ \widehat{\mathcal{L}}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \right] \nonumber \\
    &= \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \frac{1}{S}\sum_{s=1}^{S} \log \left( w_\thetab(\xb,\zb_{s}) \right) \right] \nonumber \\
    &\leq \log \mathbb{E}_{\zb_{s}\sim q_\phib(\zb|\xb)}\left[ \frac{1}{S}\sum_{s=1}^{S} w_\thetab(\xb,\zb_{s}) \right] \nonumber \\
    &= \log p(\xb) \enspace .
\end{align}
%
From this and the law of large numbers, it can be shown that for all $S$,
%
\begin{equation}
    \log p(\xb) \geq \mathcal{L}^{\text{IWAE}}_{S+1}(\xb;\thetab,\phib) \geq \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \enspace ,
\end{equation}
%
and $\lim_{S\rightarrow\infty} \mathcal{L}^{\text{IWAE}}_{S}(\xb;\thetab,\phib) \rightarrow \log p(\xb)$, if $w_\thetab(\xb,\zb_{s})$ are bounded. 
Hence, $\widehat{\mathcal{L}}^{\text{IWAE}}_{S}$ is a biased estimator of $\log p(\xb)$ but with a bias that goes to zero as the number of samples goes to infinity. 
With $S=1$, the estimator is equivalent to the regular ELBO \cref{eq:variational-autoencoder-elbo} and so, it is always as tight, or tighter, than the ELBO \parencite{burda_importance_2016}.
